---
title: "Empirical estimation of event rates and rate ratios"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Empirical estimation of event rates and rate ratios}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
has_robincar <- requireNamespace("RobinCar", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = has_robincar
)
```

```{r setup, eval = TRUE}
library(EmpiricalRate)
```

## The problem

Many clinical trials collect **count** endpoints — for example the number of
hypoglycemia episodes a participant experiences over their follow-up. The usual
summary is a *rate ratio* between arms. Negative binomial (NB) regression is the
standard tool, but its maximum-likelihood estimation can be numerically unstable
when events are sparse, and covariate adjustment can lose efficiency if the
assumed relationship between covariates and the outcome is wrong.

The empirical method targets the **same estimand** as NB regression (the ratio
of marginal event rates) without fitting a parametric count model.

## The idea in one line

For subject $j$ in arm $i$ with count $Y_{ij}$ and exposure $d_{ij}$, transform

$$W_{ij} = \frac{Y_{ij}}{\bar d_{i\cdot}}, \qquad
  \bar d_{i\cdot} = \text{mean exposure in arm } i.$$

Since $E[Y_{ij}] = d_{ij} r_i$, the arm-specific mean of $W_{ij}$ is *exactly
unbiased* for the marginal rate $r_i$ — no distributional assumption required.
Fitting a linear model to $W$ and back-transforming gives rate ratios with
confidence intervals.

## Example data

The package ships a **simulated** dataset, `hypo_sim`, with a sparse event
count, heterogeneous exposure, and two baseline covariates. (It is not real
trial data.)

```{r data, eval = TRUE}
data(hypo_sim)
head(hypo_sim)

# Observed aggregate rate in each arm:
with(hypo_sim, tapply(events, arm, sum) / tapply(exposure_years, arm, sum))
```

## Unadjusted analysis

With no covariates and simple randomization, the estimated marginal rate for
each arm equals the observed aggregate rate exactly, and the rate ratio equals
the raw rate ratio.

```{r unadjusted}
fit <- empirical_rate(
  hypo_sim,
  treatment = "arm",
  count     = "events",
  exposure  = "exposure_years",
  reference = "Control"
)
fit
```

We can confirm the identity numerically:

```{r identity}
agg <- with(hypo_sim, tapply(events, arm, sum) / tapply(exposure_years, arm, sum))
est <- setNames(fit$rates$rate, fit$rates$treatment)
rbind(observed = agg, empirical = est[names(agg)])
```

## Covariate-adjusted analysis

Adjustment uses a common-slope ANCOVA working model. Validity does not depend on
this model being correctly specified — the target is still the marginal rate.

```{r adjusted}
fit_adj <- empirical_rate(
  hypo_sim,
  treatment  = "arm",
  count      = "events",
  exposure   = "exposure_years",
  covariates = c("baseline_hba1c", "baseline_rate"),
  reference  = "Control"
)
summary(fit_adj)
```

## Accounting for a stratified randomization scheme

If the trial used stratified (covariate-adaptive) randomization, the variance
estimator can exploit it via the ANHECOVA extension. Set `car_scheme` and give
the stratification columns in `strata`. (The `hypo_sim` data were not generated
with stratified randomization, so this is shown only for syntax.)

```{r anhecova, eval = FALSE}
empirical_rate(
  hypo_sim,
  treatment  = "arm",
  count      = "events",
  exposure   = "exposure_years",
  covariates = c("baseline_hba1c", "baseline_rate"),
  car_scheme = "permuted-block",
  strata     = "your_strata_column",
  reference  = "Control"
)
```

## Pooling across strata or studies

When the rate ratio is assumed common across strata but rates differ,
`meta_rate_ratio()` pools per-stratum estimates with exposure weights. It
consumes the `log_rr` and `se_log_rr` columns from `empirical_rate()`.

```{r meta, eval = TRUE}
meta_rate_ratio(
  log_rr    = c(0.41, 0.55, 0.29),
  se_log_rr = c(0.18, 0.26, 0.15),
  weight    = c(320, 150, 410)
)
```

Pooling defaults to the log scale (recommended in the paper's Discussion);
`scale = "identity"` reproduces the natural-scale formula in Appendix A.

```{r meta-identity, eval = TRUE}
meta_rate_ratio(
  log_rr    = c(0.41, 0.55, 0.29),
  se_log_rr = c(0.18, 0.26, 0.15),
  weight    = c(320, 150, 410),
  scale     = "identity"
)
```
